Short-term synaptic plasticity and threshold adaptation act in synergy to give rise to deviance detection in a spiking model of neuronal culture

### 2022-12-07 paper-1/manuscript.tex
Rev sec Methods.Model

Our research is based on loose approximation of a matured culture of dissociated neurons. We modeled neurons as leaky integrate-and-fire units whose membrane potential $v$ followed
\begin{equation}
    \tau_m \frac{dv}{dt} = (v_{rest}-v) + I_{syn}
\end{equation}
where
\begin{align}
    I_{syn} &= g_e(E_e-v) + g_i(E_i-v) \nonumber \\
    \tau_e \frac{dg_e}{dt} &= -g_e + \sum_{j \in \boldsymbol E} U x_j w \delta(t - \hat{t_j}) \nonumber \\
    \tau_i \frac{dg_i}{dt} &= -g_i + \sum_{j \in \boldsymbol I} w \delta(t - \hat{t_j}) \label{eqn-syn}
\end{align}
with $\boldsymbol E$ and $\boldsymbol I$ the sets of presynaptic excitatory and inhibitory neurons, respectively, $w$ the synaptic weight, $\delta(\cdot)$ the Dirac delta function, and $\hat{t}$ the time of spikes.

Excitatory, but not inhibitory, synapses were subject to short-term depression (STD), simplified from \cite{Tsodyks1997-qt}; since synaptic transmission was not stochastic, we modeled the depression variable $x_j$ as a property of the presynaptic neuron $j$,
\begin{equation}
    \tau_x \frac{dx_j}{dt} = (1-x_j) - U x_j \delta(t - \hat{t})
\end{equation}
In ablated experiments without STD, $x$ remained constant at $1$, equivalent to $\tau_x = 0$, but $U$ in Equation \ref{eqn-syn} was left unchanged for balance purposes.

When a neuron's membrane potential reached $v \geq v_\theta$, a spike was emitted, and the potential was clamped to $v = v_{reset}$ for a fixed refractory period of 3~ms (excitatory neurons) or 2~ms (inhibitory neurons). For inhibitory neurons, the firing threshold remained constant at $v_\theta = \theta_0$, while for excitatory neurons, the threshold included an adaptive component, following
\begin{align}
    v_\theta &= \theta_0 + \theta \nonumber \\
    \tau_{TA} \frac{dV_{TA}}{dt} &= -V_{TA} + \hat{\theta} \delta(t - \hat{t})
\end{align}
In ablated experiments without TA, $\hat{\theta}$ was set to $0$ to abolish adaptation.

\begin{figure*}%
    \centering
    \includegraphics{fig1.pdf}
    \caption{\textbf{a} Sample network layout, showing excitatory and inhibitory neurons and all outgoing synaptic connections from a small subset. \textbf{b} Distance from stimulation site 0 in terms of minimum number of excitatory synapses. Stimulation is provided to the 10 neurons nearest to each of 5 stimulation sites, which are regularly spaced on a circle with a radius of 2.5 mm. \textbf{c} Membrane and synaptic dynamics of a neuron innervated by an excitatory and an inhibitory neuron firing pre-determined spikes. The excitatory connection has weight $w = 3$ for demonstration purposes only; a single connection with $w = 1$ is normally unable to evoke postsynaptic firing. Top: Membrane voltage (black, $\tau$ = 30 ms), firing threshold baseline (gray) and adaptive threshold (green, $\tau$ = 1 s) demonstrating TA. Middle: Excitatory ($g_e$, $\tau$ = 2 ms) and inhibitory ($g_i$, $\tau$ = 4 ms) synaptic conductances, demonstrating excitatory STD. Bottom: Inputs to $g_e$ and $g_i$ (vertical bars), and the depression variable $x$ of the excitatory presynaptic neuron (dashed line)}
    \label{fig1}
\end{figure*}

\textbf{Fig. 1 a} Sample network layout, showing excitatory and inhibitory neurons and all outgoing synaptic connections from a small subset. \textbf{b} Distance from stimulation site 0 in terms of minimum number of excitatory synapses. Stimulation is provided to the 10 neurons nearest to each of 5 stimulation sites, which are regularly spaced on a circle with a radius of 2.5 mm. \textbf{c} Membrane and synaptic dynamics of a neuron innervated by an excitatory and an inhibitory neuron firing pre-determined spikes. The excitatory connection has weight $w = 3$ for demonstration purposes only; a single connection with $w = 1$ is normally unable to evoke postsynaptic firing. Top: Membrane voltage (black, $\tau$ = 30 ms), firing threshold baseline (gray) and adaptive threshold (green, $\tau$ = 1 s) demonstrating TA. Middle: Excitatory ($g_e$, $\tau$ = 2 ms) and inhibitory ($g_i$, $\tau$ = 4 ms) synaptic conductances, demonstrating excitatory STD. Bottom: Inputs to $g_e$ and $g_i$ (vertical bars), and the depression variable $x$ of the excitatory presynaptic neuron (dashed line)

Figure \ref{fig1}c shows the internal and presynaptic dynamics of an excitatory neuron in a contrived example for the purpose of illustration. The model was implemented in Brian2 \citep{Stimberg2019-tc}, accelerated with Brian2GeNN \citep{Stimberg2020-go}, and simulated with an integration time step of 1~ms. Synaptic delay was not modeled explicitly, but spikes were delivered, and voltages reset, in the time step following spike emission. We simulated neurons without any stochasticity in either their inputs or their parameters, other than network structure (see below), in order to ensure that our results were driven purely by the experimental paradigm, rather than any model-internal sources of noise.

800 excitatory and 200 inhibitory neurons were placed randomly in a two-dimensional circular space with a radius of 4~mm, mimicking a petri dish, and formed synapses of weight $w=1$ with 50 randomly selected postsynaptic partners within a range of 2~mm (excitatory) or 1~mm (inhibitory). Connectivity and weights remained fixed throughout simulation. Stimulation sites were evenly distributed 2.5~mm from the center of the dish, and stimulation delivered a one-time increase in $g_e$ to the 10 closest neurons, sufficient to trigger 2-3 spikes each. See Figure \ref{fig1}a,b for a representative example of the resulting network structure. Networks were pseudo-randomly generated according to the above scheme and screened for minimal stimulus response, discarding networks where less than half of all neurons in the fully recovered network responded to stimulation at any of the five sites.

>>>>

We modeled neurons as leaky integrate-and-fire units whose membrane potential $v$ at time $t$ followed
\begin{equation}
    \tau_m \frac{dv}{dt} = (v_{rest}-v) + I_{syn}(t)
\end{equation}
with $tau_m = 30$~ms the membrane time constant and $v_{rest} = -60$~mV the resting membrane potential. Synaptic currents were modeled in a conductance-based manner, following
\begin{equation}
    I_{syn} &= g_e(E_e-v) + g_i(E_i-v)
\end{equation}
with $E_e = 0$~mV and $E_i = -100$~mV the excitatory and inhibitory reversal potentials, respectively. Synaptic conductances evolved according to
\begin{align} 
    \tau_e \frac{dg_e}{dt} &= -g_e + \sum_{j \in \boldsymbol E} U x_j w \delta(t - \hat{t_j}) \nonumber \\
    \tau_i \frac{dg_i}{dt} &= -g_i + \sum_{j \in \boldsymbol I} w \delta(t - \hat{t_j}) \label{eqn-gsyn}
\end{align}
with $\tau_e = 2$~ms the excitatory time constant, echoing AMPA receptor dynamics \citep{Hausser1997-cn}, $\tau_i = 4$~ms the inhibitory time constant, echoing GABA-A receptor dynamics \citep{Destexhe1994-oc}, $\boldsymbol E$ and $\boldsymbol I$ the sets of presynaptic excitatory and inhibitory neurons, respectively, $w$ the synaptic weight, $\delta(\cdot)$ the Dirac delta function, and $\hat{t}$ the spike times.

Excitatory, but not inhibitory, synapses were subject to short-term depression (STD), simplified from \cite{Tsodyks1997-qt}; since synaptic transmission was not stochastic, we modeled the depression variable $x_j$ as a property of the presynaptic neuron $j$,
\begin{equation}
    \tau_x \frac{dx_j}{dt} = (1-x_j) - U x_j \delta(t - \hat{t}) \label{eqn-xsyn}
\end{equation}
with recovery time constant $\tau_x = 150$~ms and release fraction $U = 0.4$.

When a neuron's membrane potential reached the firing threshold $v_\theta$, a spike was emitted, and the potential was clamped to $v = v_{reset} = -74$~mV for a fixed refractory period of 3~ms (excitatory neurons). Inhibitory neurons were modeled as fast-spiking cells with a refractory period of 2~ms and a constant firing threshold $v_\theta = \theta_0 = -54$~mV \refp{Mensi2012-au}. In contrast, excitatory neurons were modeled with threshold adaptation (TA) as in \cite{Teeter2018-iz}, following
\begin{align}
    v_\theta &= \theta_0 + V_{TA}(t) \nonumber \\
    \tau_{\theta} \frac{dV_{TA}}{dt} &= -V_{TA} + \hat{\theta} \delta(t - \hat{t}) \label{eqn-TA}
\end{align}
with increment $\hat{\theta} = 1$~mV and decay time constant $\tau_{\theta} = 1$~s \refp{Pozzorini2015-ei}.

\begin{figure*}%
    \centering
    \includegraphics{fig1.pdf}
    \caption{}
    \label{fig1}
\end{figure*}

\textbf{Fig. 1}Model and paradigm. \textbf{a} Membrane and synaptic dynamics of a neuron innervated by an excitatory and an inhibitory neuron, illustrated above the plots, firing pre-determined spikes. The excitatory connection has weight $w = 3$ for demonstration purposes only; a single connection with $w = 1$ is normally unable to evoke postsynaptic firing. Top: Membrane voltage (black), firing threshold baseline (gray) and adaptive threshold (green) demonstrating TA. Middle: Excitatory ($g_e$, blue) and inhibitory ($g_i$, red) synaptic conductances, demonstrating excitatory STD. Bottom: Inputs to $g_e$ and $g_i$ (vertical bars in the appropriate colors), and the depression variable $x$ of the excitatory presynaptic neuron (dashed line). \textbf{b} Sample network layout, showing excitatory and inhibitory neurons and all outgoing synaptic connections from two neurons of each type. \textbf{c} Distance from stimulation site 0 in terms of minimum number of excitatory synapses. Stimulation is provided to the 10 neurons nearest to the center of each of 5 stimulation sites (round markers), which are regularly spaced on a circle with a radius of 2.5 mm.

Figure \ref{fig1}a shows the internal and presynaptic dynamics of an excitatory neuron in a contrived example for the purpose of illustration. The model was implemented in Brian2 \citep{Stimberg2019-tc}, accelerated with Brian2GeNN \citep{Stimberg2020-go}, and simulated with an integration time step of 1~ms. Synaptic delay was not modeled explicitly, but spikes were delivered, and voltages reset, in the time step following spike emission. We simulated neurons without any stochasticity in either their inputs or their parameters, other than network structure (described below), in order to ensure that our results were driven purely by the experimental paradigm, rather than any model-internal sources of noise.

800 excitatory and 200 inhibitory neurons were placed randomly in a two-dimensional circular space with a radius of 4~mm, and formed synapses of weight $w=1$ with 50 randomly selected postsynaptic partners within a range of 2~mm (excitatory) or 1~mm (inhibitory), reflecting the notion that excitatory neurons project over longer distances, while inhibitory neurons mainly connect to their local neighborhood. Connectivity and weights remained fixed throughout. Stimulation sites were evenly distributed 2.5~mm from the center of the dish, and stimulation delivered a one-time increase in $g_e$ to the 10 closest neurons, sufficient to trigger 2-3 spikes each. See Figure \ref{fig1}b and c for a representative example of the resulting network structure. Networks were pseudo-randomly generated according to the above scheme and screened for minimal stimulus response: Networks where fewer than 500 neurons responded to stimulation at any of the five sites after full recovery were discarded. In total, 30 networks were used for the experiments described below.
